---
title: "Environmental Plastic Waste Analysis"
author: Jade Gehrts
format: html
embed-resources: true
---

```{r}
#| label: install-load-packages

library(httr)
library(tidyjson)
library(jsonlite)
library(tidyverse)
library(glue)
library(purrr)
library(countrycode)
library(scales)
```
## Context

The Environmental Plastic Waste Data comes from the **"Break Free from Plastic"** movement and covers information on plastic use across the world, organized by country, type of plastic, source of plastic, and year (2019 or 2020). The data comes from brand audits and cleanups conducted by volunteers. Some limitations of this dataset are that the countries will not have an equal sampling effort and the data depends on volunteer participation. Additionally, the amount of collected waste does not necessarily tell us how much plastic there is in general. Aside from those, the data is useful to help identify major corporate contributors to plastic pollution in our environment. It can also help inform environmental policy workers and provide corporate accountability. 

# Checkpoint 3

Checkpoint 3

Temperature (dataset): https://www.kaggle.com/datasets/palinatx/mean-temperature-for-countries-by-year-2014-2022

GDP (API):
https://www.imf.org/external/datamapper/api/help
https://api-ninjas.com/api/gdp 

Key: 
8FzJfy67lKE4OUN74J48zZqI0x2ivVWkBPIRqN0K

https://api.api-ninjas.com/v1/gdp?country=Canada%20;X-Api-Key=RID1ERp0hNnGs3eiSIxGI87z2tT3Y8FDbngGnwlb


Quarto Code:




```{r}
#| label: write-csv-gdp
#| eval: false
write_csv(x = gdp, file = "stat431-gdp")
```

The original GDP data set includes historical, current, and predicted data on GDP for countries around the world for the years for 1980-2029. We are particularly interested in the years 2019 and 2020. The 3-letter ISO code for each country is also included. There are 5 variables measuring GDP in different ways. For example, the variable `gdp_per_capita_nominal` measures GDP by dividing the country's total value of finished goods and services by its total population. `gdp_nominal` is the nominal GDP in billions of US dollars and `gdp_growth` expresses the GDP growth rate as a percentage. 


```{r}
#| label: read-csv-gdp
gdp <- read_csv("stat431-gdp.csv")
gdp
```



Checkpoint 4

# Project Checkpoint 4



```{r}
#| label: read-in-meta-data
#| echo: true
#| output: false
plastic_temp_gdp <- read_csv("stat431-meta")
```

```{r}
#| label: add-region-col
plastic_temp_gdp <- plastic_temp_gdp |>
  mutate(continent = countrycode(Country,
                                  origin = "country.name",
                                  destination = "continent"))
```

## RQ #1: How does change in GDP relate to change in the total amount of plastic collected?

Checkpoint 4

# Project Checkpoint 4


```{r}
#| label: read-in-meta- data
#| echo: true
#| output: false
plastic_temp_gdp <- read_csv("stat431-meta")
```

```{r}
#| label: add-region- col
plastic_temp_gdp <- plastic_temp_gdp |>
  mutate(continent = countrycode(Country,
                                  origin = "country.name",
                                  destination = "continent"))
```

## RQ #1: How does change in GDP relate to change in the total amount of plastic collected?

```{r}
changed <- plastic_temp_gdp |>
  group_by(Country, continent, Year) |>
  summarise(grand_total = sum(grand_total, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = Year, values_from = grand_total) |>
  rename(
    total_2019 = `2019`,
    total_2020 = `2020`
  ) |>
  mutate(change = total_2020 - total_2019) |>
  drop_na(change, continent)
```

```{r}
#| fig-cap: Figure 1.1. Barplot for Change in Grand Total Plastic from 2019-2020, by Continent
#| fig-alt: Figure 1.1 Barplot showing changes in grand total plastic collection across continents between 2019 to 2020.
continent_year <- plastic_temp_gdp |>
  group_by(continent, Year) |>
  summarise(grand_total = sum(grand_total, na.rm = TRUE), .groups = "drop")

ggplot(continent_year, aes(x = continent, y = grand_total, fill = factor(Year))) +
  geom_col(position = "dodge") +
  labs(
    title = "Grand Total by Continent (2019 vs 2020)",
    x = "Continent",
    y = "Grand Total",
    fill = "Year"
  ) +
  scale_fill_manual(values = c("steelblue", "lightgreen")) +
  theme_classic() + 
  theme(plot.title = element_text(face = "bold"))
```

```{r}
#| fig-cap: Figure 1.2. Boxplot for Change in Grand Total Plastic from 2019-2020, by Continent
#| fig-alt: Figure 1.2. Boxplot showing change in grand total plastic collection across continents between 2019 and 2020. 
ggplot(changed, aes(x = continent, y = change)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.5) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Change in Grand Total by Continent",
    subtitle = "Country-level change from 2019 to 2020",
    x = "Continent",
    y = "Change (2020 - 2019)"
  ) +
  theme_classic() + 
  theme(plot.title = element_text(face = "bold"))
```

We noticed an outlier in Africa that was that was heavily compressing the scale of the boxplot. We identified the outlier to be Nigeria. 

```{r}
#| fig-cap: Figure 1.3. Boxplot for Change in Grand Total Plastic from 2019-2020, by Continent (excluding Africa outlier, Nigeria)
#| fig-alt: Figure 1.2. Boxplot showing change in grand total plastic collection across continents between 2019 and 2020, with Nigeria (outlier) removed.
ggplot(changed, aes(x = continent, y = change)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.5) +
  coord_cartesian(ylim = c(-30000, 30000))+
  scale_y_continuous(labels = comma) +
  labs(
    title = "Change in Grand Total by Continent",
    subtitle = "Country-level change from 2019 to 2020, excluding Nigeria",
    x = "Continent",
    y = "Change (2020 - 2019)"
  ) +
  theme_classic() + 
  theme(plot.title = element_text(face = "bold"))
```

```{r}
#| fig-cap: Figure 1.4. Barplot for Change in GDP from 2019-2020, by Continent
#| fig-alt: Figure 1.4. Barplot showing change in GDP across continent from 2019 to 2020.
continent_year <- plastic_temp_gdp |>
  group_by(continent, Year) |>
  summarise(gdp_nominal = sum(gdp_nominal, na.rm = TRUE), .groups = "drop")

ggplot(continent_year, aes(x = continent, y = gdp_nominal, fill = factor(Year))) +
  geom_col(position = "dodge") +
  labs(
    title = "Nominal GDP by Continent (2019 vs 2020)",
    x = "Continent",
    y = "Nominal GDP",
    fill = "Year"
  ) +
  scale_fill_manual(values = c("steelblue", "lightgreen")) +
  theme_classic() + 
  theme(plot.title = element_text(face = "bold"))
```

```{r}
# Calculating Percent Change in GDP from 2019 to 2020
gdp_changed <- plastic_temp_gdp %>%
  arrange(Country, Year) %>%
  group_by(Country) %>%
  mutate(gdp_pct_change = (gdp_nominal / lag(gdp_nominal) - 1) * 100)
```

```{r}
#| fig-cap: Figure 1.5. Boxplot for Percent Change in GDP from 2019-2020, by Continent
#| fig-alt: Figure 1.5. Boxplot showing percent change in GDP across continents from 2019 to 2020.
#| warning: false
ggplot(gdp_changed, aes(x = continent, y = gdp_pct_change)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +
  labs(y = "GDP % Change (2019-2020)", x = "Continent") + 
  theme_classic() + 
  theme(plot.title = element_text(face = "bold"))
```


## RQ #2: Is there a relationship between a country's mean annual temperature and the total amount of plastic collected?

```{r}
#| message: false 
#| fig-cap: Figure 2. Hexbin plot of Mean Annual Temperature and Grand Total Plastic Collected (2019-2020) by Continent
#| fig-alt: Figure 2. Hexbin plot showing mean annual temperature and grand total plastic collected from 2019 to 2020, by continent.
ggplot(data = plastic_temp_gdp) +
  geom_hex(mapping = aes(x = AnnualMeanTemp, y = grand_total)) +
  scale_fill_gradient(low = "lightblue", high = "royalblue4") +
  geom_smooth(aes(x = AnnualMeanTemp, y = grand_total), 
              method = "lm", se = FALSE, 
              color = "black", linewidth = 0.5, lty = 2) +
  labs(x = "Mean Annual Temperature (C°)", 
       y = "Grand Total Plastic Collected", 
       fill = "Number of Countries",
       title = "Number of Countries Observed", 
       subtitle = "For Mean Annual Temperature, Grand Total Plastic Collected Combination (2019 - 2020)") +
  theme_classic() + 
  theme(plot.title = element_text(face = "bold")) 
```

```{r}
# Fitting a linear model
plastic.fit <- lm(grand_total~AnnualMeanTemp, data = plastic_temp_gdp)
summary(plastic.fit)
```



SHRAVANI’S CODE: 

```{r}

changed <- plastic_temp_gdp |>
  group_by(Country, continent, Year) |>
  summarise(grand_total = sum(grand_total, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = Year, values_from = grand_total) |>
  rename(
    total_2019 = `2019`,
    total_2020 = `2020`
  ) |>
  mutate(change = total_2020 - total_2019) |>
  drop_na(change, continent)
```


Barplot for Continents by Change in Grand Total Plastic from 2019-2020
```{r}
continent_year <- plastic_temp_gdp |>
  group_by(continent, Year) |>
  summarise(grand_total = sum(grand_total, na.rm = TRUE), .groups = "drop")

ggplot(continent_year, aes(x = continent, y = grand_total, fill = factor(Year))) +
  geom_col(position = "dodge") +
  labs(
    title = "Grand Total by Continent (2019 vs 2020)",
    x = "Continent",
    y = "Grand Total",
    fill = "Year"
  ) +
  scale_fill_manual(values = c("steelblue", "lightgreen")) +
  theme_minimal()
```

Boxplot for Continents by Change in Grand Total Plastic from 2019-2020

```{r}
ggplot(changed, aes(x = continent, y = change)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.5) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Change in Grand Total by Continent",
    subtitle = "Country-level change from 2019 to 2020",
    x = "Continent",
    y = "Change (2020 - 2019)"
  ) +
  theme_minimal()
```

Boxplot for Continents by Change in Grand Total Plastic from 2019-2020 (zoomed in without Africa outlier)
```{r}
ggplot(changed, aes(x = continent, y = change)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.5) +
  coord_cartesian(ylim = c(-30000, 30000))+
  scale_y_continuous(labels = comma) +
  labs(
    title = "Change in Grand Total by Continent",
    subtitle = "Country-level change from 2019 to 2020",
    x = "Continent",
    y = "Change (2020 - 2019)"
  ) +
  theme_minimal()
```
Interesting Change

- what was the outlier in Africa? Nigeria


Barplot for Continents by Change in GDP from 2019-2020
```{r}
continent_year <- plastic_temp_gdp |>
  group_by(continent, Year) |>
  summarise(gdp_nominal = sum(gdp_nominal, na.rm = TRUE), .groups = "drop")

ggplot(continent_year, aes(x = continent, y = gdp_nominal, fill = factor(Year))) +
  geom_col(position = "dodge") +
  labs(
    title = "Nominal GDP by Continent (2019 vs 2020)",
    x = "Continent",
    y = "Nominal GDP",
    fill = "Year"
  ) +
  scale_fill_manual(values = c("steelblue", "lightgreen")) +
  theme_minimal()
```

Calculating Percent Change in GDP from 2019 to 2020
```{r}
gdp_changed <- plastic_temp_gdp %>%
  arrange(Country, Year) %>%
  group_by(Country) %>%
  mutate(gdp_pct_change = (gdp_nominal / lag(gdp_nominal) - 1) * 100)
```

Boxplot for Continents by Percent Change in GDP from 2019-2020
```{r}
ggplot(gdp_changed, aes(x = continent, y = gdp_pct_change)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +
  labs(y = "GDP % Change (2019-2020)", x = "Continent") + 
  theme_minimal()
```





NOT USED


Table
```{r}
plastic_temp_gdp |>
  filter(Year == 2019) |>
  group_by(continent) |>
  summarise(total_events = sum(num_events, na.rm = TRUE)) |>
  mutate(percent = total_events / sum(total_events) * 100)
```

```{r}
plastic_temp_gdp |>
  filter(Year == 2020) |>
  group_by(continent) |>
  summarise(total_events = sum(num_events, na.rm = TRUE)) |>
  mutate(percent = total_events / sum(total_events) * 100)
```


```{r}
plastic_temp_gdp |>
  filter(Year == 2019) |>
  group_by(continent) |>
  summarise(grand_total = sum(grand_total, na.rm = TRUE), total_gdp = sum(gdp_per_capita_nominal, na.rm = TRUE)) |>
  mutate(percent = grand_total / sum(grand_total) * 100)
```
```{r}
#| fig-cap: “”
#| fig-alt: “”

rq1_long <- plastic_temp_gdp %>%
  filter(Year %in% c(2019, 2020)) %>%
  pivot_longer(
    cols = c(hdpe, ldpe, o, pet, pp, ps, pvc),
    names_to = "plastic_type",
    values_to = "amount"
  )
rq1_summary <- rq1_long %>%
  group_by(continent, plastic_type) %>%
  summarise(
    total_amount = sum(amount, na.rm = TRUE),
    .groups = "drop"
  )

rq1_prop_by_type <- rq1_summary %>%
  group_by(plastic_type) %>%
  mutate(prop = total_amount / sum(total_amount)) %>%
  ungroup()

rq1_wide_by_type <- rq1_prop_by_type %>%
  select(continent, plastic_type, prop) %>%
  pivot_wider(
    names_from = plastic_type,
    values_from = prop
  )

```


```{r}
plastic_distribution_summary <- function(data, plastic_var, plastic_name) {
  
  data |>
    mutate(
      plastic_proportion = {{ plastic_var }} / grand_total
    ) |>
    group_by(continent) |>
    summarize(
      mean_proportion = mean(plastic_proportion, na.rm = TRUE),
      median_proportion = median(plastic_proportion, na.rm = TRUE),
      total_amount = sum({{ plastic_var }}, na.rm = TRUE),
      .groups = "drop"
    ) |>
    arrange(desc(mean_proportion)) |>
    gt() |>
    tab_header(
      title = paste(plastic_name, "Distribution Across Continents"),
      subtitle = "Proportion of selected plastic type relative to total plastic collected"
    ) |>
    cols_label(
      continent = "Continent",
      mean_proportion = "Mean %",
      median_proportion = "Median %",
      total_amount = "Total Count"
    ) |>
    fmt_percent(
      columns = c(mean_proportion, median_proportion),
      decimals = 1
    ) |>
    fmt_number(
      columns = total_amount,
      decimals = 0
    )
}

```




```{r}
#| fig-cap: “Figure 3.  ”
#| fig-alt: “Figure 3. ”

continent_plastic <- plastic_long %>%
  group_by(continent, plastic_type) %>%
  summarise(total_amount = sum(amount, na.rm = TRUE), .groups = "drop")

ggplot(continent_plastic, aes(x = continent, y = total_amount, fill = plastic_type)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Plastic Type Composition by Continent",
    x = "Continent",
    y = "Proportion",
    fill = "Plastic Type"
  ) +
  theme_classic()
```

```






